Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2021-01-19 journal paper/content/related_work.tex
Edits for spacing.

\xhdr{Exploration.} Multi-armed bandits (\emph{MAB}) is an elegant and tractable abstraction for tradeoff between \emph{exploration} and \emph{exploitation}: essentially, between acquisition and usage of information. MAB problems have been studied for many decades by researchers from computer science, operations research, statistics and economics, generating a vast and multi-threaded literature.  The most relevant thread concerns the basic model of regret-minimizating bandits with stochastic rewards and no auxiliary structure (which is the problem faced by each principal in our model), see Appendix~\ref{app:bg} for background. This basic model has been extended in many different directions, with a considerable amount of work on each: \eg payoffs with a specific structure (\eg combinatorial, linear, convex or Lipschitz), payoff distributions that change over time, and auxiliary payoff-relevant signals.
%\footnote{There is a superficial similarity, in name only, between this paper and the work on ``dueling bandits" \citep[starting from][]{Yue-dueling12,Yue-dueling-icml09}. In this work, there is only one bandit algorithm which chooses two arms in each round, and the only observes which arm has ``won the duel".}
Dedicated monographs \citep{Bubeck-survey12,slivkins-MABbook,LS19bandit-book} cover the work on regret-minimizing formulations (which mainly comes from computer science). The classic book
\citep{Gittins-book11} focuses on the Markovian formulations, which predate regret-minimization. Various connections to economics and game theory are detailed in
books \citep{CesaBL-book,slivkins-MABbook} and surveys \citep{Bergemann-survey06,Horner-survey16}. Industrial applications are discussed in \citep{DS-arxiv}.

The three-way tradeoff between exploration, exploitation and incentives has been studied in several scenarios very different from ours:
incentivized exploration in recommendation systems
(starting from \citet{Kremer-JPE14,Che-13}, see \citet[Ch. 11]{slivkins-MABbook}),
    %\citep[\eg][]{Che-13,Frazier-ec14,Kremer-JPE14,ICexploration-ec15,Bimpikis-exploration-ms17,Bahar-ec16,Jieming-unbiased18},
dynamic auctions
    (\eg see \citet{DynAuctions-survey10} for a survey),
    %\citep[\eg][]{AtheySegal-econometrica13,DynPivot-econometrica10,Kakade-pivot-or13},
pay-per-click ad auctions
    \citep[\eg][]{MechMAB-ec09,DevanurK09},
coordinating search and matching
    \citep{Bobby-Glen-ec16},
and human computation
    \citep[\eg][]{RepeatedPA-ec14,Ghosh-itcs13}.
%A literature review of this work can be found in
%\citep[Ch. 11.6]{slivkins-MABbook}.

>>>>

\xhdr{Exploration.} Multi-armed bandits (\emph{MAB}) is an elegant and tractable abstraction for tradeoff between \emph{exploration} and \emph{exploitation}: essentially, between acquisition and usage of information. MAB problems have been studied for many decades by researchers from computer science, operations research, statistics and economics, generating a vast and multi-threaded literature.  The most relevant thread concerns the basic model of regret-minimizating bandits with stochastic rewards and no auxiliary structure (which is the problem faced by each principal in our model), see Appendix~\ref{app:bg} for background. This basic model has been extended in many different directions, with a considerable amount of work on each: \eg payoffs with a specific structure (\eg combinatorial, linear, convex or Lipschitz), payoff distributions that change over time, and auxiliary payoff-relevant signals.
%\footnote{There is a superficial similarity, in name only, between this paper and the work on ``dueling bandits" \citep[starting from][]{Yue-dueling12,Yue-dueling-icml09}. In this work, there is only one bandit algorithm which chooses two arms in each round, and the only observes which arm has ``won the duel".}
Dedicated monographs \citep{Bubeck-survey12,slivkins-MABbook,LS19bandit-book} cover the work on regret-minimizing formulations (which mainly comes from computer science). The classic book
\citep{Gittins-book11} focuses on the Markovian formulations, which predate regret-minimization. Connections to economics are detailed in
books \citep{CesaBL-book,slivkins-MABbook} and surveys \citep{Bergemann-survey06,Horner-survey16}. Industrial applications are discussed in \citep{DS-arxiv}.

The three-way tradeoff between exploration, exploitation and incentives has been studied in several scenarios very different from ours:
incentivized exploration in recommendation systems
(starting from \citet{Kremer-JPE14,Che-13}, see \citet[Ch. 11]{slivkins-MABbook}),
    %\citep[\eg][]{Che-13,Frazier-ec14,Kremer-JPE14,ICexploration-ec15,Bimpikis-exploration-ms17,Bahar-ec16,Jieming-unbiased18},
dynamic auctions
    \citep{DynAuctions-survey10},
    %\citep[\eg][]{AtheySegal-econometrica13,DynPivot-econometrica10,Kakade-pivot-or13},
pay-per-click ad auctions
    \citep[\eg][]{MechMAB-ec09,DevanurK09},
coordinating search and matching
    \citep{Bobby-Glen-ec16},
and human computation
    \citep[\eg][]{RepeatedPA-ec14,Ghosh-itcs13}.
%A literature review of this work can be found in
%\citep[Ch. 11.6]{slivkins-MABbook}.

### 2021-01-19 journal paper/content/sec-intro.tex
Edits for spacing.

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better. %\citep[\eg][]{aghion2005competition,Vives-08}.
We interpret innovation as the adoption of better exploration algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere \citep[\eg see][]{DS-arxiv}.}
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.
%\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different inverted-U relationship which zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.

While traditional models of innovation study lab-based R\&D, we consider data-driven innovation: one that crucially depends on data generated by the firm's customers.
We focus on innovation in \emph{exploration technology} which systematically improves the firm's products, whereas prior work would define innovation as improvement in the products themselves. We recover the traditional inverted-U relationship purely through the reputational consequences of exploration.
\asdelete{rather than monetary costs/benefits thereof.}
This is in contrast to the inverted-U relationships established in prior work, which rely on the monetary aspects: how much can a firm invest into  R\&D and how much can it profit from the innovation.

@@

\textbf{(ii)} Focusing on \HardMax, we investigate the first-mover advantage as a different channel to vary the intensity of competition: from the first-mover to simultaneous arrival to late-arriver. We find that the first-mover is incentivized to choose a more advanced exploration algorithm, whereas the late-arriver is often incentivized to choose the ``greedy algorithm" (more so than under simultaneous arrival). Consumer welfare is higher under early/late arrival than under simultaneous entry. We frame these results in terms of an inverted-U relationship.

>>>>

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better. %\citep[\eg][]{aghion2005competition,Vives-08}.
We interpret innovation as the adoption of better exploration algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere \citep[\eg see][]{DS-arxiv}.}
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.
%\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
Another, technically different inverted-U relationship zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.

While traditional models of innovation study lab-based R\&D, we consider data-driven innovation, which crucially depends on data generated by the firm's customers.
We focus on innovation in \emph{exploration technology} which systematically improves the firm's products, whereas prior work would define innovation as improvement in the products themselves. We recover the \asdelete{traditional} inverted-U relationship purely through the reputational consequences of exploration.
\asdelete{rather than monetary costs/benefits thereof.}
By contrast, the inverted-U relationships from prior work rely on the monetary aspects: \asedit{investments into  R\&D and profits from innovation.}

%how much can a firm invest into  R\&D and how much can it profit from the innovation.

@@

\textbf{(ii)} Focusing on \HardMax, we investigate the first-mover advantage as a different channel to vary the intensity of competition: from the first-mover to simultaneous entry to late-arriver. We find that the first-mover is incentivized to choose a more advanced exploration algorithm, whereas the late-arriver is often incentivized to choose the ``greedy algorithm" (more so than under simultaneous entry). Consumer welfare is higher under early/late arrival than under simultaneous entry. We frame these results in terms of an inverted-U relationship.
